---
title: "Logical variables and Filters"
author: Heike Hofmann
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```


## Comparisons


Define vector `a` to be `a <- c(1,15, 3,20, 5,8,9,10, 1,3)`

Find the expression for the logical vector that is TRUE where `a` is:

- less than 20

```{r one, exercise=TRUE}
a <- c(1,15, 3,20, 5,8,9,10, 1,3)

```

```{r one-solution}
a <- c(1,15, 3,20, 5,8,9,10, 1,3)
a < 20
```

- squared value is at least 100 or less than 10

```{r two, exercise=TRUE}
a <- c(1,15, 3,20, 5,8,9,10, 1,3)

```

```{r two-solution}
a <- c(1,15, 3,20, 5,8,9,10, 1,3)
(a^2 >= 100) | (a < 10)
```

- equals 1 or 3

```{r three, exercise=TRUE}
a <- c(1,15, 3,20, 5,8,9,10, 1,3)

```
```{r three-solution}
a <- c(1,15, 3,20, 5,8,9,10, 1,3)
a %in% c(1,3)
```

- even

```{r four, exercise=TRUE}
a <- c(1,15, 3,20, 5,8,9,10, 1,3)

```
```{r four-hint}
have a look at the help for the operator `%%`
```
```{r four-solution}
(a %% 2) == 0
```


## Plotting subsets

Get a subset of all crimes in Iowa, Plot incidences/rates for one type of crime over time

```{r iowa, exercise = TRUE, exercise.lines = 6, exercise.eval=FALSE}
data(fbi, package="classdata")
library(ggplot2)
library(dplyr)



```

```{r iowa-solution}
iowa <- filter(fbi, State=="Iowa", Type=="Larceny.theft")
ggplot(iowa, aes(x = Year, y=Count)) + geom_point()
```


Get a subset of all crimes in 2009. Plot the number or rate for one type of crime by state.

```{r cr2009, exercise = TRUE, exercise.lines = 5}
data(fbi, package="classdata")
library(ggplot2)
library(dplyr)



```

```{r cr2009-solution}
cr2009 <- filter(fbi, Year==2009, Type=="Larceny.theft")
ggplot(cr2009, aes(x = State, y=Count)) + geom_point() + coord_flip()
```

Get a subset of the data that includes number of homicides for the last five years. Find the rate of homicides, extract all states that have a rate of greater than 90% of the rates across all states, and plot (Hint: `?quantile`)

```{r murder, exercise = TRUE, exercise.lines = 10}
data(fbi, package="classdata")
library(ggplot2)
library(dplyr)



```

```{r murder-hint}
?quantile
```

```{r murder-solution}
murder <- filter(fbi, Year>=2010, Type=="Murder.and.nonnegligent.Manslaughter")
murder$Rate <- murder$Count/murder$Population*100000

# top 10%:
quantile(murder$Rate, probs=0.9)

top10 <- filter(murder, Rate >= quantile(murder$Rate, probs=0.9))
ggplot(top10, aes(x = Rate, y=Year)) + geom_point() + facet_grid(State~.)
```


<span style="color:white">Extra credit (1 point): submit your code (regardless of whether it works or not) in Canvas (yourturn-checkin-1).


## Useful Info Functions


Use the `fbi` data object to answer the following questions: 
- how many reports of Burglaries are there for California?

```{r burglaries, exercise = TRUE, exercise.lines = 10}
data(fbi, package="classdata")
library(ggplot2)
library(dplyr)



```

```{r burglaries-solution}
ca_burg <- filter(fbi, Type=="Burglary", State=="California")
nrow(ca_burg)

sum(ca_burg$Count)
```

- for any of the violent crimes, which state had the highest rate (and for which crime) in 2014? in 1961? 

```{r violent, exercise = TRUE, exercise.lines = 10}
data(fbi, package="classdata")
library(ggplot2)
library(dplyr)



```

```{r violent-solution}
violent <- filter(fbi, Violent.Crime == TRUE)
violent$Rate <- violent$Count/violent$Population*100000
  
vio2014 <- filter(violent, Year==2014)
vio2014[vio2014$Rate==max(vio2014$Rate),]

vio1961 <- filter(violent, Year==1961)
vio1961[which.max(vio1961$Rate),]

```


Use the `fbiwide` data  to answer the following questions: 

In how many states in 2011 were there more vehicle thefts than robberies? 

```{r vehicle, exercise = TRUE, exercise.lines = 10}
data(fbiwide, package="classdata")
library(ggplot2)
library(dplyr)



```

```{r vehicle-solution}
compare <- filter(fbiwide, Motor.vehicle.theft>Robbery)
compare11 <- filter(compare, Year == 2011)

unique(compare11$State)
length(unique(compare11$State)) # almost all, so where were fewer?

filter(fbiwide, Motor.vehicle.theft<=Robbery, Year==2011)
```




## Going back to the beginning


The gapminder data we originally worked with is available [here](https://stat480-at-isu.github.io/materials/01_collaborative-environment/data/gapminder-5060.csv). 

`read.csv` helps you read the gapminder from the given link. Store the result in an object:

```{r gp-read, exercise = TRUE}
```

```{r gp-read-solution}
gapminder <- read.csv("https://stat480-at-isu.github.io/materials/01_collaborative-environment/data/gapminder-5060.csv")
```

What format does the object have?

```{r gp-format, exercise = TRUE}
gapminder <- read.csv("https://stat480-at-isu.github.io/materials/01_collaborative-environment/data/gapminder-5060.csv")
```

```{r gp-format-solution}
str(gapminder)
```

Replace the problematic value in life expectancy for Canada in 1957  by 69.96. 

```{r gp-fix, exercise = TRUE}
gapminder <- read.csv("https://stat480-at-isu.github.io/materials/01_collaborative-environment/data/gapminder-5060.csv")
```

```{r gp-fix-solution}
# check that we have the right record:
gapminder[gapminder$year==1957 & gapminder$country=="Canada",] 

gapminder$lifeExp[gapminder$year==1957 & gapminder$country=="Canada"] <- 69.96
```


