---
title: "Stat 480: Factors"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```






##  Factors in Data Frames


Inspect the `fbi` object. How many variables are there? Which type does each of the variables have?

```{r ncols, exercise=TRUE}
data(fbi, package = "classdata")


```

```{r ncols-solution}
ncol(fbi)
str(fbi)
```


Make a summary of Year.

```{r year, exercise=TRUE}
data(fbi, package = "classdata")

```

```{r year-solution}
summary(fbi$Year)
```


Make Year  a factor variable, then do a summary of Year. How is this different from the previous summary?

```{r factor, exercise=TRUE}
data(fbi, package = "classdata")

```

```{r factor-solution}
fbi$Year <- factor(fbi$Year)
summary(fbi$Year)
```

Are there other variables in the `fbi` data that should be factors (or vice versa)? 

```{r fbi, exercise = TRUE}
data(fbi, package = "classdata")
fbi$Year <- factor(Year)
str(fbi)
```

```{r fbi-solution}
# Besides Year, Type is a factor variable
# State and Abb could also be considered factors
# Count and Population are observed variables.
# Violent could be turned into a factor variable, but then we lose the logical information
```



## Reorder Factor Levels

To answer the next questions, use the `fbi` object from the `classdata` package. 

Introduce a rate of the number of reported offenses by population into the `fbi` data. You could use the *Ames standard* to make values comparable to a city of the size of Ames (population ~70,000). Plot boxplots of crime rates by different types of crime. How can you make axis text legible?

```{r rate, exercise = TRUE}
library(ggplot2)
data(fbi, package="classdata")


```

```{r rate-solution}
fbi$Rate <- fbi$Count/fbi$Population * 70000

ggplot(data=fbi, aes(x = Type, y = Rate)) + geom_boxplot() +
  coord_flip()
```


Reorder the boxplots of crime rates, such that the boxplots are ordered by their medians. 

```{r reorder, exercise = TRUE}
library(ggplot2)
data(fbi, package="classdata")
fbi$Rate <- fbi$Count/fbi$Population * 70000


```

```{r reorder-solution}
ggplot(data=fbi, aes(x = reorder(Type, Rate, median, na.rm=TRUE),
                     y = Rate)) + 
  geom_boxplot() +
  coord_flip()
```


For one type of crime (subset!) plot boxplots of rates by state, reorder boxplots by median crime rates</span>


```{r subset, exercise = TRUE}
library(ggplot2)
library(dplyr)
data(fbi, package="classdata")
fbi$Rate <- fbi$Count/fbi$Population * 70000


```

```{r subset-solution}
onetype <- filter(fbi, Type=="Burglary")
ggplot(data=onetype, 
       aes(x = reorder(State, Rate, median, na.rm=TRUE),
           y = Rate)) + 
  geom_boxplot() +
  coord_flip()
```



## Changing Levels

Check on the level names given for the type of crimes in the `fbi` data:

```{r levels, exercise=TRUE}
data(fbi, package="classdata")

```

```{r levels-solution}
levels(fbi$Type)
```


Change the level for murders to `Murder`:

```{r change, exercise=TRUE}
data(fbi, package="classdata")

```

```{r change-solution}
levels(fbi$Type)[6] <- "Murder"
```

## Visualizing factor variables


For this exercise  use the `Titanic` object from the `classdata` package.


Draw a barchart of Sex. Interpret.

```{r gender, exercise=TRUE}
data(Titanic, package="classdata")
library(ggplot2)

```

```{r gender-solution}
ggplot(data=Titanic, aes(x = Sex)) + geom_bar()
# more men than women were on board the Titanic
```


Map survival to fill color in the barchart of Gender. Interpret.

```{r survived, exercise=TRUE}
data(Titanic, package="classdata")
library(ggplot2)
ggplot(data=Titanic, aes(x = Sex)) + geom_bar()

```

```{r survived-solution}
ggplot(data=Titanic, aes(x = Sex, fill=Survived)) + geom_bar()
# about as many men survived as women
```

In the previous barchart change the `position` parameter to `"fill"`. Interpret.

```{r ratio, exercise=TRUE}
data(Titanic, package="classdata")
library(ggplot2)
ggplot(data=Titanic, aes(x = Sex, fill=Survived)) + geom_bar()

```

```{r ratio-solution}
ggplot(data=Titanic, aes(x = Sex, fill=Survived)) + geom_bar(position="fill")
# the ratio of survival among women was much higher than among men.
```


Read up on the `position` parameter in `?geom_bar`. Try out other options for `position`. Interpret.

```{r dodge, exercise=TRUE}
data(Titanic, package="classdata")
library(ggplot2)

ggplot(data=Titanic, aes(x = Sex, fill=Survived)) + geom_bar()
```

```{r dodge-solution}
ggplot(data=Titanic, aes(x = Sex, fill=Survived)) + geom_bar(position="dodge")
# dodge places categories side by side. We can compare subgroups directly.
```

